How is experience transformed into memory?

### 2018-08-28 paper/main.tex
legend

What does it mean to \textit{remember} something? In traditional
episodic memory experiments \citep[e.g., list-learning or trial-based
experiments;][]{Murd62a, Kaha96}, remembering is often cast as a
discrete and binary operation: each studied item may be separated from
the rest of one's experiences, and that item may be labeled as having
been recalled versus forgotten. More nuanced studies might incorporate
self-reported confidence measures as a proxy for memory strength, or
ask participants to discriminate between ``recollecting'' the
(contextual) details of an experience or having a general feeling of
``familiarity'' \citep{Yone02}. However, characterizing and evaluating
memory in more realistic contexts (e.g., recounting a recent
experience to a friend) is fundamentally different.  First, real-world
recall is continuous, rather than discrete.  Removing an event from
the context in which it occurrs often subtantially changes its
meaning.  Second, the specific words used to describe an experience
have very little bearing on whether the experience is considered to
have been ``remembered.''  Asking whether the rememberer has precisely
reproduced a specific set of words to describe a given experience is
nearly orthogonal to whether they were actually able to remember it.
Third, one might remember the gist or essence of an experience but
forget (or neglect to recount) particular details.  Capturing the gist
of what happened is typically the ``point'' of recounting a memory to
a listener but, depending on the circumstances, accurate recall of any
specific detail may be irrelevant.

@@

While our experiences often change gradually, they can also sometimes change suddenly (e.g., walking through a doorway) \citep{RadvZack17}. Prior research suggests that these sharp transitions (coined in the literature as ``event boundaries'') during an experience may serve to descretize our memories into ``events'' \citep{RadvZack17, BrunEtal18, HeusEtal18, ClewDava17, EzzyDava11, DuBrDava13}. Short-timescale autocorrelations in the contents of an experience are thought to associatively link information in memory, while more drastic changes in perceptual or semantic aspects of an experience may lead to discontinuities in memory \citep{HeusEtal18, BrunEtal18, EzzyDava11, DuBrDava13}.  In other words, the interplay between stable and transient temporal dynamics may impose structure on memories for those experiences. There is now compelling evidence that event boundaries can influence memory organization when learning sequences of items \citep{HeusEtal18, DuBrDava13}, during navigation \citep{BrunEtal18}, and narrative comprehension \citep{ZwaaRadv98, EzzyDava11}.  However, whether people verbally recount their experiences in a way that is structured around events has not been established.

@@

\subsubsection*{Modeling the naturalistic stimulus and verbal recall}
We fit a topic model \citep{BleiEtal03} to hand-annotated text descriptions of scenes from the video. The text descriptions contained details of the scene such as the characters present, location, and a short summary of the scene (see Fig.~\ref{fig:schematic} for example text). As depicted in Fig.~\ref{fig:model}A, the topic vectors are sparse and change slowly over time. Furthermore, there are clear transitions from one topic ``state'' to the next, possibly indexing scene transitions in the stimulus. To get a better handle on this temporal structure, we computed a timepoints (1976) by timepoints (1976) correlation matrix of the video model (Fig.~\ref{fig:model}C).  This correlation matrix reveals that the model has a strong, block-diagonal structure with very low off-diagonal correlation values, suggesting the representations of each block are unique and highly descriminable. We found that the ``narrative details'' (a few sentences summarizing the dialogue and specific happenings of the scene) of the video were the most important feature driving this model structure (see Supp. Fig.~\ref{fig:feature-importance}A).

@@

\subsubsection*{Choosing topic model parameters}
There were 3 critical parameters related to fitting the topic model: 1) the number of topics, 2) the window size of text descriptions of the video used to fit the model, and 3) the window size of recall sentences used to transform the recall data.  To chose these parameter values, we performed a grid search where the range of possible parameter values was 1, 5, 10, 25, 50, 100, 200, and 500. Our optimization objective was defined as the correlation between the hand annotated memory performance and the root mean squared distance between the video model and the recall model before any further processing (e.g., hidden Markov modelling, averaging within event, etc). While many of the parameter combination elicited moderately high correlations, the optimal choice was 100 topics, 50 video segments and 10 recall sentences.
%NOTE: I did run this analysis, but before that i arbitrarily selected some parameters that ended up being very close to the best.  The reason I bring this up is that the correlations between some of the memory metrics (k, precision, distinctiveness) and hand annotated accuracy could be biased because we searched for the optimal parameters. Its not `the same' measure, but they are related to each other, so we'll have to be extra careful not to over interpret the correlations in this study.

@@

\begin{suppfigure}[t!]
\centering
\includegraphics[width=1\textwidth]{figs/supp4_feature_value.pdf}
\caption{\small \textbf{Impact of individual features on topic modeling analysis.} \textbf{A.} Contribution of each feature to model structure. Bars represent the correlation of a video model trained in the absence of a given feature to the model trained on all features. \textbf{B.} Contribution of each feature to video model/recall model relationship. The leftmost bar represents the across-participants mean correlation between the video model and recall models trained on all features. Subsequent bars represent the same relationship between video and recall models trained in the absence of a given feature. Error bars are the standard error of the mean across participants.}
\label{fig:feature-importance}
\end{suppfigure}

>>>>

What does it mean to \textit{remember} something? In traditional
episodic memory experiments \citep[e.g., list-learning or trial-based
experiments;][]{Murd62a, Kaha96}, remembering is often cast as a
discrete and binary operation: each studied item may be separated from
the rest of one's experiences, and that item may be labeled as having
been recalled versus forgotten. More nuanced studies might incorporate
self-reported confidence measures as a proxy for memory strength, or
ask participants to discriminate between ``recollecting'' the
(contextual) details of an experience or having a general feeling of
``familiarity'' \citep{Yone02}. However, characterizing and evaluating
memory in more realistic contexts (e.g., recounting a recent
experience to a friend) is fundamentally different.  First, real-world
recall is continuous, rather than discrete.  Removing an event from
the context in which it occurs often substantially changes its
meaning.  Second, the specific words used to describe an experience
have very little bearing on whether the experience is considered to
have been ``remembered.''  Asking whether the rememberer has precisely
reproduced a specific set of words to describe a given experience is
nearly orthogonal to whether they were actually able to remember it.
Third, one might remember the gist or essence of an experience but
forget (or neglect to recount) particular details.  Capturing the gist
of what happened is typically the ``point'' of recounting a memory to
a listener but, depending on the circumstances, accurate recall of any
specific detail may be irrelevant.

@@

While our experiences often change gradually, they can also sometimes change suddenly (e.g., walking through a doorway) \citep{RadvZack17}. Prior research suggests that these sharp transitions (coined in the literature as ``event boundaries'') during an experience may serve to discretize our memories into ``events'' \citep{RadvZack17, BrunEtal18, HeusEtal18, ClewDava17, EzzyDava11, DuBrDava13}. Short-timescale autocorrelations in the contents of an experience are thought to associatively link information in memory, while more drastic changes in perceptual or semantic aspects of an experience may lead to discontinuities in memory \citep{HeusEtal18, BrunEtal18, EzzyDava11, DuBrDava13}.  In other words, the interplay between stable and transient temporal dynamics may impose structure on memories for those experiences. There is now compelling evidence that event boundaries can influence memory organization when learning sequences of items \citep{HeusEtal18, DuBrDava13}, during navigation \citep{BrunEtal18}, and narrative comprehension \citep{ZwaaRadv98, EzzyDava11}.  However, whether people verbally recount their experiences in a way that is structured around events has not been established.

@@

\subsubsection*{Modeling the naturalistic stimulus and verbal recall}
We fit a topic model \citep{BleiEtal03} to hand-annotated text descriptions of scenes from the video. The text descriptions contained details of the scene such as the characters present, location, and a short summary of the scene (see Fig.~\ref{fig:schematic} for example text). As depicted in Fig.~\ref{fig:model}A, the topic vectors are sparse and change slowly over time. Furthermore, there are clear transitions from one topic ``state'' to the next, possibly indexing scene transitions in the stimulus. To get a better handle on this temporal structure, we computed a timepoints (1976) by timepoints (1976) correlation matrix of the video model (Fig.~\ref{fig:model}C).  This correlation matrix reveals that the model has a strong, block-diagonal structure with very low off-diagonal correlation values, suggesting the representations of each block are unique and highly discriminable. We found that the ``narrative details'' (a few sentences summarizing the dialogue and specific happenings of the scene) of the video were the most important feature driving this model structure (see Supp. Fig.~\ref{fig:feature-importance}A).

@@

\subsubsection*{Choosing topic model parameters}
There were 3 critical parameters related to fitting the topic model: 1) the number of topics, 2) the window size of text descriptions of the video used to fit the model, and 3) the window size of recall sentences used to transform the recall data.  To chose these parameter values, we performed a grid search where the range of possible parameter values was 1, 5, 10, 25, 50, 100, 200, and 500. Our optimization objective was defined as the correlation between the hand annotated memory performance and the root mean squared distance between the video model and the recall model before any further processing (e.g., hidden Markov modeling, averaging within event, etc). While many of the parameter combination elicited moderately high correlations, the optimal choice was 100 topics, 50 video segments and 10 recall sentences.
%NOTE: I did run this analysis, but before that i arbitrarily selected some parameters that ended up being very close to the best.  The reason I bring this up is that the correlations between some of the memory metrics (k, precision, distinctiveness) and hand annotated accuracy could be biased because we searched for the optimal parameters. Its not `the same' measure, but they are related to each other, so we'll have to be extra careful not to over interpret the correlations in this study.

@@

\begin{suppfigure}[t!]
\centering
\includegraphics[width=1\textwidth]{figs/supp4_feature_value.pdf}
\caption{\small \textbf{Impact of individual features on topic modeling analysis.} \textbf{A.} Contribution of each feature to model structure. Bars represent the correlation of a video model trained in the absence of a given feature to the model trained on all features. \textbf{B.} Contribution of each feature to video model/recall model relationship. The leftmost bar represents the across-participants mean correlation between the video model and recall models trained on all features. Subsequent bars represent the same relationship between video and recall models trained in the absence of a given feature. Error bars are the standard error of the mean across participants. \textbf{C.}  Individual feature trajectory similarity matrix. The first row/column is the topic trajectory the full video model. Each subsequent row/column is that of a single feature. Shading corresponds to the value of the correlation coefficient (Pearson's $r$) .}
\label{fig:feature-importance}
\end{suppfigure}
